---
title: "session08 - homework"
format: html
editor: visual
date: 2025-07-17
author: Timo Roettger
execute:
  error: false
  warning: false
  message: false
  cache: true
bibliography: ../resources/bibliography.bib
---

# Preamble: Loading packages and configuration

```{r}
#| label: data_and_libraries
#| echo: false

# just run this code chunk
# function to ignoring the setting of the relative path below when knitting
run_if_not_knitting <- function(expr) {
  if (!isTRUE(getOption("knitr.in.progress"))) {
    eval(expr)
  }
}

# nifty code using the pacman package
# it checks if the packages specified below are installed, if not, they will be installed, if yes, they will be loaded
if (!require("pacman")) install.packages("pacman")
pacman::p_load(rstudioapi, tidyverse, tidybayes, brms, bayesplot, bayestestR, modelr)

# set the current working directory to the one where this file is
run_if_not_knitting(current_working_dir <- dirname(rstudioapi::getActiveDocumentContext()$path))
run_if_not_knitting(setwd(current_working_dir))

# options to increase efficiency of brms models (optional)
#options(brms.backend = "cmdstanr")
options(mc.cores = parallel::detectCores())

```

# Exercise

We will be re-analyzing the data from @winter2013making.

```{r}
#| label: preprocessing_sim

# Load our datasets
sim <- read_csv("../../data/linguistic_similarity.csv") |> 
  mutate(Sem = as.factor(Sem),
         Phon = as.factor(Phon))
sim

```

We will now retrace the original article's complete analysis. In their original analysis, they were interested in whether (a) semantic similarity between cities, (b) phonological similarity between cities, and (c) their interaction affects how close/far apart participants draw the cities.

## (a) Specify Priors

Specify the model formula and appropriate weakly informative priors for all relevant predictors. Contrast-code the predictors.

```{r}
#| label: model_formula_sim

# define model formula
sim_formula <- bf(Distance ~ Sem * Phon)

# get priors
get_prior(sim_formula, sim)

# set contrasts
contrasts(sim$Sem) <- contr.sum(2)
contrasts(sim$Phon) <- contr.sum(2)

# check that new contrasts have been applied
contrasts(sim$Sem)
contrasts(sim$Phon)

# specify weakly informative priors
my_priors <- c(prior(normal(74, 34.1), class = "Intercept"),
               prior(normal(0, 34.1), class = "b", coef = "Phon1"),
               prior(normal(0, 34.1), class = "b", coef = "Sem1"),
               prior(normal(0, 34.1), class = "b", coef = "Sem1:Phon1")
               )
```

## (b) Run the appropriate model for the research question

```{r}
#| label: model_sim

# run model
xmdl_sim <- brm(sim_formula,
                family = gaussian(),
                prior = my_priors,
                data = sim,
                chains = 4,
                cores = 4,
                iter = 2000,
                warmup = 1000,
                thin = 1,
                seed = 21
                )
```

## (c) Posterior predicive check

Check the overall model fit with posterior predictive checks. If the model does not fit the data well, fix the problem (check `06_homework_solutions.qmd` for inspiration).

```{r}
#| label: model_fit

# run post predicative check
pp_check(xmdl_sim, ndraws = 100)

# fix assumption violation with a gentle square root transform
sim <- sim |> 
  mutate(sqrt_dist = sqrt(Distance))

# get priors
get_prior(sqrt_dist ~ Sem * Phon, sim)

# specify new priors
my_priors <- c(prior(normal(8.6, 2.5), class = "Intercept"),
               prior(normal(0, 2.5), class = "b", coef = "Phon1"),
               prior(normal(0, 2.5), class = "b", coef = "Sem1"),
               prior(normal(0, 2.5), class = "b", coef = "Sem1:Phon1")
               )

# run model
xmdl_sqrt_sim <- brm(sqrt_dist ~ Sem * Phon,
                family = gaussian(),
                prior = my_priors,
                data = sim,
                chains = 4,
                cores = 4,
                iter = 2000,
                warmup = 1000,
                thin = 1,
                seed = 21
                )

# run post predicative check
pp_check(xmdl_sqrt_sim, ndraws = 100)

```

## (d) Assess sampling diagnostics

Check that chains are reasonably mixed and that sampling yielded a sufficient number of effective sample size. If there are issues, address them accordingly.

```{r}
#| label: sampling_diagnostics

# plot Markov chain Monte Carlo (MCMC) simulations
mcmc_plot(xmdl_sqrt_sim, type = "trace")
  # trace plots look good
#plot denisty estimates
mcmc_dens_overlay(xmdl_sqrt_sim, pars = c("b_Sem1", "b_Phon1", "b_Sem1:Phon1"))
  # density estimate plots look good
# check model output
xmdl_sqrt_sim
  # Rhat = 1.00 suggests that the chains have converged (i.e., proper convergence of model).
  # Bulk_ESS (4984 > 400) and Tail_ESS (3055 > 100) are both satisfactory.
```

## (e) Interpret the results

Now interpret the model relative to the research question and describe uncertainty, visualize posteriors, calculate posterior probabilities of direction, and calculate Bayes Factors for evidence for and against the posterior coefficients being point zero.

(Note: When you want to extract interaction terms like `b_Sem1:Phon1`, you have to put single quotation marks around them because R struggles with interpreting the colon)

```{r}
#| label: interpret_sim

# extract posterior draws from model
sim_draws <- xmdl_sqrt_sim |> 
  spread_draws(b_Intercept, b_Sem1, b_Phon1, `b_Sem1:Phon1`)

# calculate posterior probability
p_direction(sim_draws)
```

## (g) Write up your entire analysis:

YOUR WRITE UP HERE

## OPTIONAL (h) compare your interpretation to the original findings

YOUR COMPARISON HERE

# References

<!-- References will auto-populate in the refs div below -->

::: {#refs}
:::
